R for Clinical Study Reports and Submission

### 2022-01-20 slides/r4csr-gwu.Rmd

---
title: "R for Clinical Study Reports and Submission"
subtitle: "The China-R Conference"
author: "Yilong Zhang, Nan Xiao, Yujie Zhao, Keaven Anderson"
date: "November 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "msd"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
---

@@

@@

- We try to fill in gaps to streamline workflow using R:
  + To develop, validate, deliver analysis results.
  + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).

@@

"FDA does not require use of any specific software for statistical analyses, and statistical software
is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
11]. However, the software package(s) used for statistical analyses should be fully documented
in the submission, including version and build identification."

@@

@@

## Tools for study design

> Details will be skipped in today's talk.

- [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
- [`simtrial`](https://github.com/Merck/simtrial): an experimental R package for simulation-based group sequential design under non-proportional hazards
- [`gsDesign2`](https://github.com/Merck/gsDesign2) and
[`gsdmvn`](https://github.com/Merck/gsdmvn):
an experimental R package for analytical-based group sequential design under proportional hazards.

Bookdown: <https://keaven.github.io/gsd-deming/>

Training: [December 6th, Deming Conference 2021](https://demingconference.org/programs/2021-program/)

@@

Training: [December 2nd, Short course for ASA Princeton-Trenton Chapter](https://rutgers.webex.com/rutgers/j.php?RGID=rc0a71b2f239233a7850a121cf972045f)

@@

- The Software Development Life Cycle
  - Planning: define the scope of a project
  - Development: mplement target deliverables
  - Validation: verify target deliverables
  - Operation: deliver work to stakeholders

@@

## Join us

- Contact us for R developer or Methodology Research Positions:
  + Yilong Zhang: <https://elong0527.github.io/>
  + Nan Xiao: <https://nanx.me/>
  + Keaven Anderson: <https://www.linkedin.com/in/keaven-anderson-09aa7538/>

- Multiple open positions (statisticians/programmers) in China, EU, and US
  + US: <https://jobs.merck.com/bards>
  + China/EU: <https://jobs.msd.com/cn/zh>

@@

>>>>

---
title: "R for Clinical Study Reports and Submission"
subtitle: "The George Washington University Biostatistics Center"
author: "Yilong Zhang"
date: "Jan 2022"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "merck"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
bibliography: reference.bib  
---

@@

> "FDA does not require use of any specific software for statistical analyses, and statistical software
is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
11]. However, the software package(s) used for statistical analyses should be fully documented
in the submission, including version and build identification."

## Motivation

As an organization, we need to ensure compliance and reduce the risk of using R and R packages in regulatory deliverables.

- R is widely used in clinical trial study design.
  + Following @lachin1986evaluation
  + [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
  + [`simtrial`](https://github.com/Merck/simtrial), 
  [`gsDesign2`](https://github.com/Merck/gsDesign2) and 
  [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
  + Bookdown: <https://keaven.github.io/gsd-deming/>

- R is flexible for handling novel missing data approaches following estimand framework in ICH E9 (R1)
  + Recurrent event data: @gao2017control @diao2020efficient 
  + Robustness: @liu2021multiply

- R is used in Bayesian statistics.
  + [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis

- R is widely used in visualization 
  + [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics) 
  + [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)

@@

- We try to fill in gaps to streamline workflow using R for clinical trial development:
  + To develop, validate, and deliver analysis results.
  + To submit analysis results to regulatory agencies in [eCTD format](https://en.wikipedia.org/wiki/Electronic_common_technical_document).

@@

# R Consortium Pilot Submission

## Challenge and Assumption

### Challenge

- How to submit internal developed (proprietary) R packages? 
- How to follow [ICH and/or FDA guidances](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd) 
in preparing eCTD package?
- How to enhance reproducibility from FDA reviewer's perspective?

### Assumption

- Focus on analysis and reporting given available ADaM datasets.

## Deliverables

- Open source pilot eCTD submission package to FDA 
  + Proprietary R package
  + R scripts for analysis
  + Analysis data reviewer guide (ADRG)
  + etc. 
- Development: <https://github.com/RConsortium/submissions-pilot1> 
- eCTD package: <https://github.com/RConsortium/submissions-pilot1-to-fda> 
- [R consortium announcement](https://www.r-consortium.org/blog/2021/12/08/successful-r-based-test-package-submitted-to-fda)

## FDA response

- "Using R version 4.1.1, FDA was able to run the submitted code and confirm the applicant's tables and the submitted figure in report-tlf pdf file."
- "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
- [original FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)

## Future work

- Address minor issues reported by FDA reviewers. 
- Pilot R Shiny app submission.
- Pilot submission with advanced analysis methods (e.g. study design, missing data, Bayesian, etc).

# Reproducibility

## Reproducibility spectrum

```{r, out.width = "100%", out.height = "400px", echo = FALSE, fig.align = "center"}
knitr::include_graphics("reproducibility.png")
```

@@

## Recommendations

Although FDA did not expect submitted R code is executable, 
sponsor shall enhance reproducibility.

- Fixed R version: e.g. R4.1.0 
- Fixed R package snapshot date: e.g. 2021-08-31 
  + [RStudio package manager](https://packagemanager.rstudio.com/client/#/repos/1/overview)
  + [MRAN](https://mran.microsoft.com/timemachine)
- Flexibility of input and output path: 
  + Define path as parameter. 
- eCTD deliverables dry-run in Windows environment
- Provide steps to follow in [ADRG appendix](https://github.com/RConsortium/submissions-pilot1-to-fda/blob/main/m5/datasets/rconsortiumpilot1/analysis/adam/datasets/adrg.pdf)

# Implementation

@@

@@

@@

- The Software Development Life Cycle
  - Planning: define the scope of a project
  - Development: implement target deliverables
  - Validation: verify target deliverables
  - Operation: deliver work to stakeholders

@@

@@

## Reference
